Predator clusters

### 2023-03-14 reports/04_pred_clus.qmd
Reduce predator clusters

library(ggmosaic)
library(ggtree)
library(sf)
library(targets)
library(tidyverse)

tar_load(c("predators_stations", "predators_abundant", "sightings_clust",
           "zoop_sf"), 
         store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
```

@@

>>>>

library(ggmosaic)
library(ggtree)
library(sf)
library(targets)
library(tidyverse)
library(vegan)

tar_load(c("predators_stations", "predators_abundant", "sightings_clust",
           "sightings_mtx", "sightings_dist", "zoop_sf"), 
         store = here::here("_targets"))
source(here::here("R", "sp.R"))
source(here::here("R", "zoop.R"))
```

@@

## Optimal number of clusters

We use the gap statistic [@tibshirani2001] to choose the optimal number of clusters ($k$). The gap statistic ($f$) is a goodness of clustering measure. The authors recommended choosing the smallest $k$ such that $f(k) \geq f(k+1) - s_{k+1}$ where $s$ is the standard error of $f$. According to this rule, the predators are best described by two clusters.

```{r}
cluster_fn <- function(x, k) {
  bray_dist <- function(x) {
    vegdist(x, method = "bray")
  }
  list(cluster = cutree(hclust(bray_dist(x), method = "ward.D2"), k = k))
}
sightings_gap <- suppressWarnings(
  cluster::clusGap(sightings_mtx, 
                   FUN = cluster_fn,
                   K.max = 10, 
                   B = 500,
                   verbose = FALSE)
)
sightings_gap_df <- as_tibble(sightings_gap$Tab) %>% 
  mutate(
    k = seq_along(logW),
    is_optimal = k == cluster::maxSE(gap, SE.sim, method = "Tibs2001SEmax")
  )
firstmaxk <- with(sightings_gap_df,
                  cluster::maxSE(gap, SE.sim, method = "firstmax"))
gap_thr <- with(sightings_gap_df, gap[firstmaxk] - SE.sim[firstmaxk])
ggplot(sightings_gap_df, aes(x = k, y = gap)) +
  geom_line() +
  geom_pointrange(aes(ymin = gap - SE.sim, ymax = gap + SE.sim,
                      color = is_optimal)) +
  geom_hline(yintercept = gap_thr, color = "red", linetype = "dashed") +
  scale_color_manual(values = c(`TRUE` = "red",
                                `FALSE` = "black")) +
  theme_classic() +
  theme(legend.position = "none")
```

## 2-cluster analysis

Repeat the indicator species analysis and maps with two clusters.

```{r}
#| label: predator_dendro2

# Cut predators into 2 clusters
clust <- cutree(sightings_clust, 2)
g <- split(names(clust), clust)
p <- ggtree(sightings_clust, hang = -1)
pred_clust_mrca <- sapply(g, function(n) MRCA(p, n))

```{r}
#| label: indicator_species2

species_wide <- pivot_wider(predators_abundant,
                            names_from = species,
                            values_from = count,
                            values_fill = 0) %>% 
  column_to_rownames("amlr.station")
cluster_indval <- labdsv::indval(species_wide, clust)
indval_mtx <- cluster_indval$indval
indval_fmt <- apply(indval_mtx, 2, function(x) {
  ord <- order(x, decreasing = TRUE)
  result <- ifelse(x >= 0.25, 
         str_glue("{rownames(indval_mtx)} ({round(x * 100)}%)"),
         NA) %>% 
    `[`(ord) %>% 
    na.omit() %>% 
    paste(collapse = "<br>")
  ifelse(result == "", "None", result)
})
as_tibble(t(indval_fmt)) %>% 
  knitr::kable(format = "html", escape = FALSE) %>% 
  kableExtra::kable_styling()
```

```{r}
#| label: cluster_maps2

## NMDS

```{r}
#| label: nmds

sightings_nmds <- metaMDS(
  sightings_mtx, 
  distance = "bray",
  k = 3,
  trymax = 100
)

stressplot(sightings_nmds)

```{r}
#| label: loading

nmds_df <- sightings_nmds$points %>% 
  as_tibble(rownames = "amlr.station") %>% 
  rename_with(~ paste0("N", .x), starts_with("MDS")) %>% 
  right_join(tibble(amlr.station = names(clust),
                    pred_clust = factor(clust)),
             by = "amlr.station")

ggplot(nmds_df, aes(NMDS1, NMDS2)) +
  geom_point(aes(color = pred_clust)) +
  theme_classic()

ggplot(nmds_df, aes(NMDS1, NMDS3)) +
  geom_point(aes(color = pred_clust)) +
  theme_classic()

ggplot(nmds_df, aes(NMDS2, NMDS3)) +
  geom_point(aes(color = pred_clust)) +
  theme_classic()
```
